Ignition of Fuel Beds by Firebrands

### 2022-05-27 conclusions.tex
Updates from comments

\chapter{Conclusions \& Future Work}
\label{part:conclusion}
\label{part:results}
The preceding chapters of this document presented findings from studies evaluating the influence of various parameters on the probability of ignition of fuel beds in wild fires. These studies were conducted to address the specific objectives of this work and to add to the current body of knowledge regarding ignition with the overall goal of facilitating a model for quickly and accurately predicting ignition risk to reduce the number of homes lost to wildfires. The specific objectives of this work are as follows:
        \begin{enumerate}
            \item Determine the effects of particle morphology on ignition propensity
            \item Ascertain ignition dependence on heating location(s), mode, and rate
            \item Identify and quantify the influence of environmental conditions on ignition propensity.
            \item Identify the influence of fuel bed chemical composition on flaming ignition.
        \end{enumerate}
The conclusions of each effort are restated in the following sections followed by a discussion of the implications of the results as they pertain to each specific objective and the overall objective of this work.

@@

\section{Species}

@@

            \item At wind speeds of 5.8\si{\meter\per\second} the ignition threshold is largely independent of the heater configuration. This lack of sensitivity is attributed to the ignition being controlled by the fluid dynamics around the firebrand surrogates. Under wind ignition is largely controlled by the propensity of pyrolysis products to accumulate in recurculation zones and independent of thermal interactions with nearby objects whether they are energy sources or sinks. 
            \item At wind speeds of 0.5\si{\meter\per\second} the ignition threshold is dependent on the firebrand surrogate configuration. For example, when the firebrand were spaced one diameter apart the difference between ignition thresholds when the upstream firebrand is inert or an energy source is 34\%. When the firebrand surrogates were spaced such that they did not thermally interact (five diameters apart) the difference in ignition threshold was 4\%. Thus, at low wind speeds or in quiescent conditions ignition is sensitive to thermal interactions with nearby objects. Ignition is promoted if nearby objects supply energy and inhibited if energy sinks are nearby.

            \item In configurations where the firebrand surrogate is downstream of an inert flow obstacle ignition may occur as a smoldering to flaming transition. The smoldering to flaming transition appears to be facilitated by either accumulation of pyrolysis gases in the downstream edge of the flow obstruction or by the formation of an overhang as the fuel bed recedes underneath the obstruction. This phenomena produces ignition at similar temperatures to other configurations and may not be important for ignition predictions but may be of interest to smoldering research. 
        \end{enumerate}

\section{Overall model}
    The overall objective of this work is to identify parameters and processes that control the ignition of a fuel bed when an ember lands on it with an anticipated impact of enabling the creation of a simplified model that may be used quickly determine the likelihood of ignition. To evaluate the effectiveness of this work in meeting those goals a model was created to predict ignition. The model was then applied to results of other studies that utilized different fuel bed materials and ignition sources to determine cross study applicability.

@@

    The aggregated data was then used to create a Random Forest Classifier model using the scikit-learn package in python~\cite{scikit-learn} to estimate ignition outcomes for each test. Using a 0.25/0.75 test train spit on the data the R\textsuperscript{2} validation score for the model with an out of bag score of 0.83. Suggesting that the model accounts can predict the ignition correctly 83\% of the time. While the accuracy of the model is likely acceptable if such a model was to be used as an indicator or predictor of ignition during a fire or to inform preventative measures it would be beneficial to reduce the number of inputs required. An examination of the most important input parameters and refinement of the model found that accuracy was not degraded if number of parameters was reduced to include only the following variables:

@@

    Using the five characteristics above the model, a prediction accuracy of 83\% was maintained. While it is beneficial to have an accurate predictive model for a single set of experiments in a similar apparatus, a model capable of predictions across different is far more valuable for preventing home loss. To evaluate the accuracy of the model across studies the model was applied to a selection of studies from literature. The studies and relevant parameters from each study are shown in Table~\ref{tab:otherStudies}. All of the studies presented in Table~\ref{tab:otherStudies} used hot metal particles dropped onto fine fuel beds of processed material. The studies by Hadden et al. and Urban et al. used pure cellulose as the fuel bed and the study by Zak et al used a blend of grasses processed into a powder. A wind direction of 0\si{\degree} was used for all studies since the geometry of the metal particles is anticipated to produce recirculation zones similar to that of a cylindrical heater parallel to the flow. The average temperature was defined as the average of the initial temperature and ambient temperature. More sophisticated estimates of the average temperature based on the average time to ignition and estimated heat losses from the ember were conducted but the accuracy was not improved. Thus, a simpler metric was chosen.  
        \begin{table}[]

@@

    The first approach used was to apply reduced model created from only the data generated from studies in this dissertation and predict ignition for each of the studies in literature. The accuracy of the model at predicting ignition for other studies was 50\%, which is entirely unusable for any sort of predictive modelling. However when the model was retrained using the same five parameters but including the Zak et al. data in conjunction with the results of this work the predictive accuracy increased to 96\% for the Hadden et al. study and 81\% for the Urban et al. study. The increase in accuracy suggests that using as few as the five parameters in this model predictions can be made across differences in experimental conditions. Predictions across experimental conditions has been one of the biggest hurdles to creating a general predictive model. The identification of the primary controlling parameters of ignition in this work has created a framework for bridging the predictive gap across various configurations.

\section{Future Work}

>>>>

\chapter{Conclusions \& Future Work}
\label{chap:conclusion}
The preceding chapters of this document presented findings from studies evaluating the influence of various parameters on the probability of ignition of fuel beds in wild fires. These studies were conducted to address the specific objectives of this work and to add to the current body of knowledge regarding ignition with the overall goal of facilitating a model for quickly and accurately predicting ignition risk to reduce the number of homes lost to wildfires. The specific objectives of this work are as follows:
        \begin{enumerate}
            \item Determine the effects of particle morphology on ignition propensity
            \item Ascertain ignition dependence on heating location(s), mode, and rate
            \item Identify and quantify the influence of environmental conditions on ignition propensity.
            \item Identify the influence of fuel bed chemical composition on flaming ignition.
        \end{enumerate}
The conclusions of each effort are restated in the following sections followed by a discussion of the implications of the results as they pertain to each specific objective and the overall objective of this work.

@@

\section{Species}
        \begin{enumerate}
            \item Of the six materials tested, only three were able to be ignited within below the 750\si{\celsius} maximum temperature of the firebrand surrogate. Two of the three materials for which ignition was not observed were bark. The higher ignition threshold of the bark materials is attributed to higher lignin concentrations. The resistance to ignition present in these materials suggest that using materials high in lignin in locations at risk for firebrand attack in wildfires may reduce ignitions and subsequent losses of structures.

            \item For Douglas-fir and pine wood, an increase in wind speed decreased the temperature required for ignition. However, for oak wood, an increase in wind speed inhibited ignition. Increases in ignition propensity with an increase in wind are attributed to the formation of recirculation zones near the ignition source. In the case of oak wood, the formation of recirculation zones inhibits ignition. The root cause of the dichotomy in ignition trends is unclear; however, differences in the flammability of pyrolysis products corresponding to changes in chemical composition are most likely.

            \item A direct correlation between the fuel bed composition and the ignition propensity was not observed. This is attributed to the complexity of biomass pyrolysis. However, a local minimum ignition threshold near the composition of Douglas-fir wood is present that warrants further investigation. 
        \end{enumerate}
    The conclusions of this work show that ignition is sensitive to the chemical composition of fuel beds in both quiescent conditions and under wind. An increase in wind was observed to increase ignition probability for some materials; however, this was not universally the case. Materials high in lignin were among the least likely to ignite and more suitable for placement in areas where firebrands are likely to land or accumulate in a WUI environment. Further and more targeted analysis of the differences in pyrolysis products for each material will likely yield further insight.

@@

            \item At wind speeds of 5.8\si{\meter\per\second} the ignition threshold is largely independent of the heater configuration. This lack of sensitivity is attributed to the ignition being controlled by the fluid dynamics around the firebrand surrogates. Under wind, ignition is largely controlled by the propensity of pyrolysis products to accumulate in recirculation zones and independent of thermal interactions with nearby objects, whether they are energy sources or sinks. 
            \item At wind speeds of 0.5\si{\meter\per\second} the ignition threshold is dependent on the firebrand surrogate configuration. For example, when the firebrands were spaced one diameter apart, the difference between ignition thresholds when the upstream firebrand is inert or an energy source is 34\%. When the firebrand surrogates were spaced such that they did not thermally interact (five diameters apart), the difference in ignition threshold was 4\%. Thus, at low wind speeds or in quiescent conditions, ignition is sensitive to thermal interactions with nearby objects. Ignition is promoted if nearby objects supply energy and inhibited if energy sinks are nearby.

            \item In configurations where the firebrand surrogate is downstream of an inert flow obstacle, ignition may occur as a smoldering to flaming transition. The smoldering to flaming transition is facilitated by either accumulation of pyrolysis gases in the downstream edge of the flow obstruction or by forming an overhang as the fuel bed recedes underneath the obstruction. This phenomenon produces ignition at similar temperatures to other configurations and may not be important for ignition predictions but may be of interest to smoldering research. 
        \end{enumerate}
    The conclusions of this work show that ignition of a fuel bed is more likely when a firebrand or firebrand lands on a fuel bed under conditions that promote recirculation zones near locations where pyrolysis gases are released from the fuel bed. It was observed that under windy conditions ignition is less sensitive to nearby embers or inert objects than at low wind speeds. These observed sensitivities highlight the increased risk of fuel bed ignition in windy conditions and in cases where firebrands may accumulate in all wind conditions.

\section{Overall model}
    The overall objective of this work is to identify parameters and processes that control the ignition of a fuel bed when an ember lands on it with an anticipated impact of enabling the creation of a simplified model that may be used to determine the likelihood of ignition quickly. A model was created to predict ignition to evaluate the effectiveness of this work in meeting those goals. The model was then applied to the results of other studies that utilized different fuel bed materials and ignition sources to determine cross-study applicability.

@@

    The aggregated data was then used to create a Random Forest Classifier model using the scikit-learn package in python~\cite{scikit-learn} to estimate ignition outcomes for each test. Using a 0.25/0.75 test train spit on the data, the R\textsuperscript{2} validation score for the model with an out-of-bag score of 0.83, which means that the model predicts ignition correctly 83\% of the time. While the accuracy of the model is likely acceptable if such a model was to be used as an indicator or predictor of ignition during a fire or to inform preventative measures, it would be beneficial to reduce the number of inputs required. An examination of the most important input parameters and refinement of the model found that accuracy was not degraded if the number of parameters was reduced to include only the following variables:

@@

    Reducing the model to include only the five characteristics above maintained the 83\% prediction accuracy. While it is beneficial to have an accurate predictive model for a single set of experiments in a similar apparatus, a model capable of predictions across different is far more valuable for preventing home loss. The model was applied to select studies from the literature to evaluate the accuracy across studies. The authors and relevant parameters from each study are shown in Table~\ref{tab:otherStudies}. All of the studies presented in Table~\ref{tab:otherStudies} used hot metal particles dropped onto fine fuel beds of processed material. The studies by Hadden et al. and Urban et al. used pure cellulose as the fuel bed, and the study by Zak et al. used a blend of grasses processed into a powder. A wind direction of 0\si{\degree} was used for all studies since the geometry of the metal particles is anticipated to produce recirculation zones similar to that of a cylindrical heater parallel to the flow. The average temperature was defined as the average of the initial temperature and ambient temperature. More sophisticated estimates of the average temperature based on the average time to ignition and estimated heat losses from the ember were conducted, but the model accuracy was not improved. Thus, a simpler metric was chosen.  
        \begin{table}[hbpt]

@@

    The first approach used was to apply a reduced model created from only the data generated from studies in this dissertation and predict ignition for each of the studies in the literature. The accuracy of the model at predicting ignition for other studies was 50\%, which is entirely unusable for any type of predictive modeling. However, when the model was retrained using the same five parameters but including the Zak et al. data in conjunction with the results of this work, the predictive accuracy increased to 96\% for the Hadden et al. study and 81\% for the Urban et al. study. The increase in accuracy suggests that predictions can be made across differences in experimental conditions using as few as five parameters. Predictions across experimental conditions have been one of the biggest hurdles to creating a general predictive model. Identifying the primary controlling parameters of ignition in this work has created a framework for bridging the predictive gap across various configurations.

\section{Future Work}
  The results in this work have identified controlling parameters for ignition and provided a framework for creating models with capabilities to predict various ignition scenarios. Significant additional work is needed to refine the proposed framework and explore different ignition parameters. 
        \begin{itemize}
            \item The moisture content of the fuel bed was not considered in this work but is a well-studied parameter that significantly affects ignition. Extension of the experiments is necessary to determine how moisture content fits into the proposed framework. The change in fuel bed bulk density due to moisture may sufficiently characterize the difference, but that remains to be seen.

            \item Additional tests with well-characterized smoldering or flaming embers would significantly enhance the knowledge gained in this work. Typical temperatures of physically combusting firebrands are higher than those capable by the apparatus used in this work~\cite{Fateev2017a}. The higher temperatures may influence the ignition propensity. However, combusting embers are susceptible to heat losses and environmental factors where the firebrand surrogates used in this work are mainly independent of these factors. It is unclear how or if the addition of energy sources that react with the environment will shift the parameters identified to be most influential to ignition. A series of highly instrumented and controlled experiments focused on the coupled interactions between a reacting ember, a reacting fuel bed, and changing environmental conditions. I applaud the patience and perseverance of those who may undertake this endeavor in the future.

            \item Relatively limited data is available regarding the thermal properties of porous biomass media. The lack of available information is likely due to the significant variation in solid biomass material properties and the endless configurations (e.g., porosity and particle orientation) in the natural environment. A series of targeted studies determining the thermal properties of common fuel bed materials would likely increase the understanding of how ignition changes between different materials or even similar materials under various packing conditions.

            \item In addition to the limited data available for thermal properties of fuel bed materials, characterization of the chemical composition of materials and the influence of chemical changes on the pyrolysis and subsequent ignition would provide valuable insight into the differences in ignition between materials and environmental conditions. 
        \end{itemize}

### 2022-05-27 introduction.tex
Updates from comments

\chapter{Introduction}
\label{part:intro}
    Wildfires are an inevitable and often beneficial occurrence in many ecosystems around the globe. However, wildfires are also a significant source of destruction, especially when they transition from ecosystems that are adapted to them to the built environment where homes, structures, and sometimes whole communities are consumed. In recent years the number of homes and structures lost to wildfires has increased and is likely to further increase due to three factors. Climate change is currently and is anticipated to be a driving factor for increased fire severity~\cite{Levin2021Unveiling2019/2020}. Wildfire exclusion has led to significant changes in ecosystems, promoting more severe fires~\cite{Marlon2012Long-termUSA, Keeley2019} and the expansion of the Wildland Urban Interface (WUI) has increased the number of homes in the path of fires~\cite{Radeloff2018RapidRisk, Hammer2009DemographicManagement, }. While the risk of loss due to fires is likely to increase for the foreseeable future, adapting structures to the increased risk by accurately predicting risks both before and during fires may reduce the number of structures lost~\cite{Manzello2021}.

   One significant mode of structure ignition is by firebrands~\cite{Manzello2020}. A firebrand, alternatively called an ember, is a hot combusting particle of biomass that travels from an active fire to another surface or biomass (e.g., pine needles, leaves, or a crevice in a deck). If the firebrands have sufficient energy, they may ignite the material they land on, leading to ignition, spot fires, and potentially the destruction of a structure. Characterization of firebrands typically involves three parts: the generation of a firebrand in the fire itself, transport of a firebrand from the fire to the eventual landing point, and the interaction of the firebrand with the surrounding material~\cite{Babrauskas2003}.

    This work aims to identify parameters and processes that control the ignition of a fuel bed when an ember lands on it. Four specific objectives were addressed to act as a framework for achieving the overall goal based on the current knowledge of fuel bed ignition. The specific objectives of this work are as follows:
        \begin{enumerate}
            \item Determine the effects of sizes of the recipient fuels on ignition behavior
            \item Ascertain ignition dependence on heating location(s), mode, and rate
            \item Identify and quantify the influence of wind on ignition propensity.
            \item Identify the influence of fuel bed chemical composition on flaming ignition.
        \end{enumerate}
    The structure of this dissertation is as follows. First, the current state of knowledge of fuel bed ignition modeling is summarized as it applies to this work (Chapter~\ref{part:literature}). A literature review specific to each objective is contained in the corresponding manuscripts. The results of this effort are then presented in manuscript form, followed by the conclusions, suggestions for future work, and appendices. The first manuscript addresses sensitivities of ignition to the particle size of the fuel bed materials (Chapter~\ref{part:manuscript1}). The second manuscript explores the influence of wind and fluid phenomena around the embers in the presence of a heat source unaffected by wind (Chapter~\ref{part:manuscript2}). The third manuscript identifies sensitivities to ignition related to chemical composition and thermal properties in both quiescent conditions and with wind (Chapter~\ref{part:manuscript3}). The fourth manuscript identifies changes in ignition propensity when multiple embers are located close to the fuel bed (Chapter~\ref{part:manuscript1}). The final chapter summarizes the conclusions of each manuscript, discusses the implications of the conclusions with respect to predictive using an example ignition model, and suggests avenues for future work (Chapter~\ref{part:conclusion}).

    It is anticipated that this work will provide insights and a basis for the creation of a simplified model that can accurately predict ignition. For example, in a recent conversation with a developer of the Fire Dynamics Simulator (FDS) the need was expressed for an accurate and low computational cost methodology or model to predict ignition for better fire spread predictions. Due to the large scales in fire simulations (~10\si{\meter} grid size) a detailed model of ignition (<1\si{\milli\meter} grid size) is not feasible. Similarly, risk management personnel or homeowners are not likely to have the capability to create detailed models and predictions of risk around structures and homes they are tasked with protecting. The goal and perceived impact of this work is to create or enable the creation of models that can be effectively implemented by the people interacting with fires and their risks in the field.

>>>>

\chapter{Introduction}
\label{chap:intro}
    Wildfires are an inevitable and are often a  beneficial occurrence in many ecosystems around the globe. However, wildfires are also a significant source of destruction, especially when they transition from ecosystems adapted to them to the built environment where homes, structures, and sometimes whole communities are consumed. In recent years the number of homes and structures lost to wildfires has increased and is likely to further increase due to three factors. Climate change is currently and is anticipated to be a driving factor for increased fire severity~\cite{Levin2021Unveiling2019/2020}. Wildfire exclusion has led to significant changes in ecosystems, promoting more severe fires~\cite{Marlon2012Long-termUSA, Keeley2019} and the expansion of the Wildland Urban Interface (WUI) has increased the number of homes in the path of fires~\cite{Radeloff2018RapidRisk, Hammer2009DemographicManagement, }. While the risk of loss due to fires is likely to increase for the foreseeable future, adapting structures to the increased risk by accurately predicting risks both before and during fires may reduce the number of structures lost~\cite{Manzello2021}.

   A significant mode of structure ignition is by firebrands~\cite{Manzello2020}. A firebrand, alternatively called an ember, is a hot combusting particle of biomass that travels from an active fire to another surface or biomass (e.g., pine needles, leaves, or a crevice in a deck). If the firebrands have sufficient energy, they may ignite the material they land on, leading to ignition, spot fires, and potentially the destruction of a structure. Characterization of firebrand and spot fire processes typically involve three parts: the generation of a firebrand in the fire itself, transport of a firebrand from the fire to the eventual landing point, and the interaction of the firebrand with the surrounding material~\cite{Babrauskas2003}.

    The overall goal of this work is to identify parameters and processes that control the ignition of a fuel bed when an ember lands on it. Four specific objectives were addressed to provide a framework for achieving the overall goal based on the current knowledge of fuel bed ignition. The specific objectives of this work are as follows:
        \begin{enumerate}
            \item Determine the effects of sizes of the recipient fuels on ignition behavior,
            \item Ascertain ignition dependence on heating location(s), mode, and rate,
            \item Identify and quantify the influence of wind on ignition propensity,
            \item Identify the influence of fuel bed chemical composition on flaming ignition
        \end{enumerate}
    The structure of this dissertation is as follows. First, the current state of knowledge of fuel bed ignition modeling is summarized as it applies to this work (Chapter~\ref{chap:literature}). A literature review specific to each objective is contained in the corresponding manuscripts. The results of this effort are then presented in manuscript form, followed by the conclusions, suggestions for future work, and appendices. The first manuscript addresses sensitivities of ignition to the particle size of the fuel bed materials (Chapter~\ref{chap:manuscript1}). The second manuscript explores the influence of wind and fluid phenomena around the embers in the presence of a heat source unaffected by wind (Chapter~\ref{chap:manuscript2}). The third manuscript identifies sensitivities to ignition related to chemical composition and thermal properties in both quiescent conditions and with wind (Chapter~\ref{chap:manuscript3}). The fourth manuscript identifies changes in ignition propensity when multiple embers are located close to the fuel bed (Chapter~\ref{chap:manuscript4}). The final chapter summarizes the conclusions of each manuscript, discusses the implications of the conclusions with respect to predictive using an example ignition model, and suggests avenues for future work (Chapter~\ref{chap:conclusion}).

    It is anticipated that this work will provide insights and a basis for the creation of a simplified model that can accurately predict ignition. For example, in a recent conversation with a developer of the Fire Dynamics Simulator (FDS), the need was expressed for an accurate and low computational cost methodology or model to predict ignition for better fire spread predictions. Due to the large scales in fire simulations (~10\si{\meter} grid size) a detailed model of ignition (<1\si{\milli\meter} grid size) is not feasible. Similarly, risk management personnel or homeowners are not likely to have the capability to create detailed models and predictions of risk around structures and homes they are tasked with protecting. The goal and perceived impact of this work is to create or enable the creation of models that can be effectively implemented by the people interacting with fires and their risks in the field.
